---
title: "Estimating the case fatality ratio for COVID-19 globally"
output: 
  html_document:
    theme: cosmo
bibliography: resources/library.bib
csl: resources/bmj.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

*Timothy W Russell<sup>1 ††</sup>, Joel Hellewell<sup>1†</sup>, Christopher I Jarvis<sup>1†</sup>, Kevin Van Zandvoort<sup>1†</sup>, Sam Abbott<sup>1</sup>, Ruwan Ratnayake<sup>1,2</sup>, CMMID COVID-19 working group, Stefan Flasche<sup>1</sup>, Rosalind M Eggo<sup>1</sup>, W John Edmunds<sup>1</sup>, Adam J Kucharski<sup>1</sup>*

*corresponding author
  

## Aim

To estimate the infection and case fatality ratio of COVID-19, using data online from the ECDC, correcting for delays between confirmation-and-death.

## Current CFR results {.tabset}

### naive CFR estimates

```{r include=FALSE}
nCFREstimates <- readRDS("../data/current_ncfr_estimates.rds")

```

```{r include = TRUE}
knitr::kable(nCFREstimates, full_width = F)
```

### corrected CFR estimates

```{r include=FALSE}
cCFREstimates <- readRDS("../data/current_ccfr_estimates.rds")

```

```{r include = TRUE}
knitr::kable(cCFREstimates, full_width = F)
```

## Adjusting for outcome delay in CFR estimates

During an outbreak, the so-called naive CFR (nCFR), i.e. the ratio of reported deaths date to reported cases to date, will underestimate the true CFR because the outcome (recovery or death) is not known for all cases [4]. We can therefore estimate the true denominator for the CFR (i.e. the number of cases with known outcomes) by accounting for the delay from confirmation-to-death [5]. 
We assumed the delay from confirmation-to-death followed the same distribution as estimated hospitalisation-to-death, based on data from the COVID-19 outbreak in Wuhan, China, between the 17th December 2019 and the 22th January 2020, accounting right-censoring in the data as a result of as-yet-unknown disease outcomes (Figure 1, panels A and B) [6]. As a sensitivity analysis, we also consider raw “non-truncated” distributions, which do not account for censoring; the raw and truncated distributions have a mean of 8.6 days and 13 days respectively.  
To correct the CFR, we use the case and death incidence data to estimate the number of cases with known outcomes [5]:

$$
u_{t} = \frac{\sum_{i = 0}^t \sum_{j = 0}^{\infty} c_{i-j} f_j}{\sum_{i = 0}^t c_i},
$$ 

where $c_{t}$ is the daily case incidence at time, $t$, $f_t$ is the proportion of cases with delay between onset or hospitalisation and death. $u_t$ represents the underestimation of the known outcomes [5] and is used to scale the value of the cumulative number of cases in the denominator in the calculation of the cCFR.

## References

1 Kucharski AJ, Edmunds WJ. Case fatality rate for ebola virus disease in west africa. The Lancet 2014;384:1260.

2 Nishiura H, Klinkenberg D, Roberts M et al. Early epidemiological assessment of the virulence of emerging infectious diseases: A case study of an influenza pandemic. PLoS One 2009;4.

3 Linton NM, Kobayashi T, Yang Y et al. Incubation period and other epidemiological characteristics of 2019 novel coronavirus infections with right truncation: A statistical analysis of publicly available case data. Journal of Clinical Medicine 2020;9:538.

4 Wu Z, McGoogan JM. Characteristics of and important lessons from the coronavirus disease 2019 (covid-19) outbreak in china: Summary of a report of 72 314 cases from the chinese center for disease control and prevention. JAMA 2020.


